Simple repression draft

### 2020-06-03 doc/section_01_introduction.tex

Gene expression presides over much of the most important dynamism of living
organisms.   The level of expression of batteries of different genes is altered
as a result of spatiotemporal cues that integrate chemical, mechanical and other
types of signals.  The original repressor-operator model conceived by Jacob and
Monod in the context of bacterial metabolism has now been transformed into the
much broader subject of gene regulatory networks in living organisms of all
kinds~\cite{Jacob1961, Britten1969, Ben-TabouDe-Leon2007}.  One of the remaining
outstanding challenges to have emerged in the genomic era is our continued
inability to predict the regulatory consequences of different regulatory
architectures, i.e. the arrangement and affinity of binding sites for
transcription factors and RNA polymerases on the DNA. This challenge stems
first and foremost from our ignorance about what those architectures even are
with more than 60\% of the genes even in an ostensibly well understood organism
such as  {\it E. coli} with no regulatory insights at all~\cite{Rydenfelt2014-2,
Belliveau2018, Ghatak2019, Santos_Zavaleta2019}. But even once we have
established the identity of key transcription factors and their binding sites of
a given promoter architecture, there remains the predictive challenge of
understanding its input-output properties, an objective that can be met by a
myriad of approaches using the tools of statistical physics~\cite{Ackers1982,
Shea1985, Buchler2003, Vilar2003a, Vilar2003b, Bintu2005a, Bintu2005c,
Gertz2009,Sherman2012, Saiz2013, Ko1991, Peccoud1995, Record1996, Kepler2001,
Sanchez2008, Shahrezaei2008, Sanchez2011, Michel2010}.  One root to such
predictive understanding is to focus on the simplest regulatory architecture and
to push the theory-experiment dialogue as far and as hard as it can be
pushed~\cite{Phillips2019}.  If we demonstrate that we can pass that test by
successfully predicting both the means and variance in gene expression, then
that provides a more solid foundation upon which to launch into more complex
problems.

To that end, in this paper we examine a wide variety of distinct models for the
simple repression regulatory architecture. This genetic architecture consists of
a DNA promoter regulated by a transcriptional repressor that binds to a single
binding site \cite{Garcia2011a}. All of the proposed models models coarse-grain
away some of the important microscopic features of this architecture that have
been elucidated by generations of geneticists, molecular biologists and
biochemists. Our aim is to understand the underlying principles of such
coarse-graining in order to increase the predictive power of these models. In
doing so we hope future models of regulatory response will be able to serve the
powerful predictive role needed to take synthetic biology from a brilliant
exercise in enlightened empiricism to a rational design framework as  in any
other branch of engineering. More precisely, we want phenomenology in the sense
of coarse-graining away atomistic detail, but still retaining biophysical
meaning. For example, we are not satisfied with the strictly phenomenological
approach offered by the commonly used Hill functions~\cite{Frank2013}.
Studies like~\cite{Razo-Mejia2018} have demonstrated that Hill functions are
clearly insufficient since each new situation requires a completely new set of
parameters. Such work requires a quantitative theory of how biophysical changes
at the molecular level propagate to input-output functions at the genetic
circuit level. We want concepts, not mere facts. In particular a key question
is: at this level of coarse graining, what microscopic details do we need to
explicitly model, and how do we figure that out? For example, do we need to
worry about all or even any of the steps that individual RNAPs go through each
time they make a transcript? Turning the question around, can we see any imprint
of those processes in the available data? If the answer is no, then those
processes are irrelevant for our purposes. Forward modeling and inverse
(statistical inferential) modeling are necessary to tackles such questions.

>>>>

Gene expression presides over much of the most important dynamism of living
organisms.   The level of expression of batteries of different genes is altered
as a result of spatiotemporal cues that integrate chemical, mechanical and other
types of signals.  The original repressor-operator model conceived by Jacob and
Monod in the context of bacterial metabolism has now been transformed into the
much broader subject of gene regulatory networks in living organisms of all
kinds~\cite{Jacob1961, Britten1969, Ben-TabouDe-Leon2007}.  One of the remaining
outstanding challenges to have emerged in the genomic era is our continued
inability to predict the regulatory consequences of different regulatory
architectures, i.e. the arrangement and affinity of binding sites for
transcription factors and RNA polymerases on the DNA. This challenge stems
first and foremost from our ignorance about what those architectures even are
with more than 60\% of the genes even in an ostensibly well understood organism
such as  {\it E. coli} with no regulatory insights at all~\cite{Rydenfelt2014-2,
Belliveau2018, Ghatak2019, Santos_Zavaleta2019}. But even once we have
established the identity of key transcription factors and their binding sites of
a given promoter architecture, there remains the predictive challenge of
understanding its input-output properties, an objective that can be met by a
myriad of approaches using the tools of statistical physics~\cite{Ackers1982,
Shea1985, Buchler2003, Vilar2003a, Vilar2003b, Bintu2005a, Bintu2005c,
Gertz2009,Sherman2012, Saiz2013, Ko1991, Peccoud1995, Record1996, Kepler2001,
Sanchez2008, Shahrezaei2008, Sanchez2011, Michel2010}.  One root to such
predictive understanding is to focus on the simplest regulatory architecture and
to push the theory-experiment dialogue as far and as hard as it can be
pushed~\cite{Garcia2011,Phillips2019}.  If we demonstrate that we can pass that test by
successfully predicting both the means and variance in gene expression, then
that provides a more solid foundation upon which to launch into more complex
problems.

To that end, in this paper we examine a wide variety of distinct models for the
simple repression regulatory architecture. This genetic architecture consists of
a DNA promoter regulated by a transcriptional repressor that binds to a single
binding site as developed in pioneering early work on
the quantitative dissection of transcription \cite{Oehler1994, Oehler1990}. All of the proposed models models coarse-grain
away some of the important microscopic features of this architecture that have
been elucidated by generations of geneticists, molecular biologists and
biochemists. Our aim is to understand the underlying principles of such
coarse-graining in order to increase the predictive power of these models. In
doing so we hope future models of regulatory response will be able to serve the
powerful predictive role needed to take synthetic biology from a brilliant
exercise in enlightened empiricism to a rational design framework as  in any
other branch of engineering. More precisely, we want phenomenology in the sense
of coarse-graining away atomistic detail, but still retaining biophysical
meaning. For example, we are not satisfied with the strictly phenomenological
approach offered by the commonly used Hill functions~\cite{Frank2013}.
Studies like~\cite{Razo-Mejia2018} have demonstrated that Hill functions are
clearly insufficient since each new situation requires a completely new set of
parameters. Such work requires a quantitative theory of how biophysical changes
at the molecular level propagate to input-output functions at the genetic
circuit level. We want concepts, not mere facts. In particular a key question
is: at this level of coarse graining, what microscopic details do we need to
explicitly model, and how do we figure that out? For example, do we need to
worry about all or even any of the steps that individual RNAPs go through each
time they make a transcript? Turning the question around, can we see any imprint
of those processes in the available data? If the answer is no, then those
processes are irrelevant for our purposes. Forward modeling and inverse
(statistical inferential) modeling are necessary to tackles such questions.
